Bow shocks, bow waves, and dust waves

### 2018-05-28 papers/Papers1234/sec-dust-bow-intro.tex
More about the rip point

\begin{figure}
  \includegraphics[width=\linewidth]{figs/test-Fdrag-components}
  \caption{Contributions of different collider species to the
    dimensionless drag force, \(f\drag / f_*\), as a function of
    gas--grain slip velocity, \(w\).  Solid lines show the Coulomb
    (electrostatic) drag, while dashed lines show the Epstein
    (solid-body) drag.  Results are shown for dimensionless grain
    potential \(\phi = 10\).  All Coulomb forces scale with
    \(\phi^2\), while the Epstein forces are independent of \(\phi\).  The
    species labelled ``CNO++'' represents the combined effect of all
    metals (total abundance \num{8.5e-4}, effective atomic weight
    \num{15.3}), which are assumed to be doubly ionized.}
  \label{fig:drag-components}
\end{figure}

@@

The drag force on a charged dust grain moving at a relative speed
\(w\) through a plasma has contributions from both direct collisions
and from electrostatic Coulomb interactions with ions and electrons.
We use the expressions in \citet{Draine:1979a}, equations~(4)--(6),
summing over protons, electrons, and helium ions.\footnote{Helium is
  assumed to be 50\% neutral and 50\% singly ionized, leading to only
  a small contribution to the drag force.  For much hotter stars, such
  as the central stars of planetary nebulae, helium may be doubly
  ionized, which leads to a more significant drag contribution,
  especially around \(w \approx \SI{5}{km.s^{-1}}\).} %
Results are shown in Figure~\ref{fig:gas-grain-drag-photoionized} for
different values of the dimensionless grain potential:
\(\phi = e^2 Z\grain / a kT\), which is the electrostatic potential
energy of a unit charge at the surface of a grain of charge
\(Z\grain\) and radius \(a\), in units of the characteristic thermal
energy of a gas particle.  The drag force \(f\drag\) is put in dimensionless units by dividing by
a characteristic force:
\begin{equation}
  \label{eq:fstar}
  f_* = 2 n k T \cdot \pi a^2 \ , 
\end{equation}
which is approximately\footnote{%
  To simplify the exposition, the gas pressure in this section is
  calculated assuming a fully ionized, pure hydrogen plasma, yielding
  \(P\gas = 2 n k T\). For typical ISM abundances, the contribution of
  helium and its corresponding electrons yield a correction to this of
  order 5\%.  The required modifications when a cool star interacts
  with a predominantly neutral gas stream are discussed later.  } %
the ionized gas pressure multiplied by the grain geometric cross
section.

For grains with low electric charge, \(\phi^2 \ll 1\), the drag force is
dominated by direct collisions of protons with the grain.  The gas
collisional mean free path is much larger than the grain size, so the
drag is in the Epstein regime \citep{Weidenschilling:1977b}.  This is
illustrated by the \(\phi = 0.25\) case (blue line) in
Figure~\ref{fig:gas-grain-drag-photoionized}. As the relative
gas--grain slip speed, \(w\), increases, \(f\drag\) first increases
linearly with \(w\) reaching \(f\drag \approx f_*\) at
\(w = \sound \approx \SI{10}{km.s^{-1}}\), then transitions to a quadratic
increase in the supersonic regime.

As \(\abs{\phi}\) increases, long-range electrostatic interactions with
protons within the Debye radius (Coulomb drag) become increasingly
important at subsonic relative velocities, as shown in
Figure~\ref{fig:gas-grain-drag-photoionized} by the orange
(\(\phi = 1\)), green (\(\phi = 4\)), and red (\(\phi = 16\)) lines.  However,
the Coulomb drag has a peak when \(w\) is equal to the thermal
speed of the colliders, which is \(\approx \SI{10}{km.s^{-1}}\) for protons,
giving a maximum strength of
\begin{equation}
  \label{eq:fdrag-maximum}
  f_{\mathrm{max}} = 0.5\, (\ln\Lambda)\, \phi^2 f_* \approx 10\, \phi^2 f_* \ , 
\end{equation}
where \(\Lambda\) is the plasma parameter (number of particles within a
Debye volume), such that
\(\ln\Lambda = 23.267 + 1.5 \ln T_4 - 0.5 \ln n\).  At highly super-thermal
speeds, the Coulomb drag falls asymptotically as
\(f\drag \propto 1 / w^{2}\).  The thermal speed of electrons is higher than
that of the protons by a factor of \((m_p / m_e)^{1/2}\), so that the
electron Coulomb drag gives a second peak fof similar strength, but at
\(w \approx \SI{430}{km.s^{-1}}\).  The behavior of \(f\drag\) in all these
different regimes is summarised in Table~\ref{tab:fdrag-regimes}, in
terms of \(\phi\) and \(w_{10} = w / \SI{10}{km.s^{-1}}\).  This is
further illustrated in Figure~\ref{fig:drag-v-phi-plane}, where the
drag regimes are located on the \((w, \abs{\phi})\) plane.

@@

From equations~\eqref{eq:dust-rad-force} and~\eqref{eq:fstar}, we can
write the radiation force acting on a grain as
\begin{equation}
  \label{eq:frad-Xi}
  f\rad = \Qp\, \Xi\, f_* \ ,
\end{equation}
where \(\Qp\) is the grain's radiation pressure efficiency (see
footnote~\ref{fn:Qp} in Appendix~\ref{sec:gas-free-bow}) and \(\Xi\) is
the local radiation parameter, defined as the ratio of direct stellar
radiation pressure to gas pressure:
\begin{equation}
  \label{eq:Xi-Prad-over-Pgas}
  \Xi \equiv \frac{P\rad}{P\gas} \approx \frac{L}{4 \pi R^2 c\, (2 n k T)} \ ,
\end{equation}
where the last expression corresponds to the optically thin limit.
The grain potential \(\phi\) is also primarily determined by \(\Xi\), as
shown in Appendix~\ref{sec:cloudy-models-dust} and the inset graph of
Figure~\ref{fig:gas-grain-drag-photoionized}, but with a slow
dependence, which can be approximated as
\begin{equation}
  \label{eq:phi-vs-Xi}
  \phi \approx 1.5 \bigl( 2.3 +  \ln \Xi \bigr) \ .
\end{equation}
There are also slight secondary dependencies on the grain composition and
stellar spectrum.  The relationship given in eq.~\eqref{eq:phi-vs-Xi}
is appropriate for graphite grains and for stellar effective
temperatures in the range \SIrange{20}{30}{kK}.  For hotter stars than
this, \(\phi\) should be multiplied by a further factor of \(1.5\), while
for silicate grains it should be divided by \(1.5\).

@@

As the dusty stream approaches the star, the radiation parameter
\(\Xi\) will increase, with a dependence of \(R^{-2}\) once the stream
is well away from the ionization front.  This increases \(f\rad\)
(eq.~\eqref{eq:frad-Xi}), but also increases the grain potential,
\(\phi\) (eq.~\eqref{eq:phi-vs-Xi}) due to the increasing dominance of
grain charging by photoelectric ejection.  Initially, this results in
a lowering of the equilibrium drift velocity to \(w_{10} \sim 0.01\) as
the Coulomb drag kicks in (see Appendix~\ref{sec:cloudy-models-dust}).
However, at smaller radii the slow logarithmic increase in
\(\phi(\Xi)\) means that the drift velocity must start increasing again to
accommodate the linear increase of \(f\rad(\Xi)\).  Eventually,
\(f\rad\) exceeds \(f_{\mathrm{max}}\), the maximum drag force that
proton Coulomb interactions can provide
(eq.~\eqref{eq:fdrag-maximum}).  This occurs at a critical value of
the radiation parameter, which we denote the \textit{rip point}:
\(\Xi_\dag \sim 1000\).  The variations in \(\Xi_\dag\) with star and grain
parameters, which are of order \SI{+- 0.5}{dex}, are listed in
Table~\ref{tab:Xi-rip}.

\begin{table}
  \caption{Critical values of radiation parameter at the rip point: \(\Xi_\dag\)}
  \centering
  \begin{tabular}{l @{\quad\quad\quad} S S} \toprule
    & \multicolumn{2}{c}{Grain composition} \\
    Spectrum & {Graphite} & {Silicate}
    \\ \midrule
    B star & 1000 +- 400 & 350 +- 150 \\
    O star & 3000 +- 500 & 2500 +- 500 \\
    \bottomrule
    \addlinespace
    \multicolumn{3}{@{}p{0.6\columnwidth}@{}}{
    Uncertainties represent variations with grain size and gas density
    (see Appendix~\ref{sec:cloudy-models-dust}).}
  \end{tabular}
  \label{tab:Xi-rip}
\end{table}

\subsection{The case of inside-out bows}
\label{sec:case-inside-out}

>>>>

\begin{figure}
  \includegraphics[width=\linewidth]{figs/test-Fdrag-components}
  \caption{Contributions of different collider species to the
    dimensionless drag force, \(f\drag / f_*\), as a function of
    gas--grain slip velocity, \(w\).  Solid lines show the Coulomb
    (electrostatic) drag, while dashed lines show the Epstein
    (solid-body) drag.  Results are shown for dimensionless grain
    potential \(\phi = 10\).  All Coulomb forces scale with
    \(\phi^2\), while the Epstein forces are independent of \(\phi\).  The
    species labelled ``CNO++'' represents the combined effect of all
    metals (see footnote~\ref{fn:metal-drag}).}
  \label{fig:drag-components}
\end{figure}

@@

The drag force on a charged dust grain moving at a relative speed
\(w\) through a plasma has contributions from both direct collisions
and from electrostatic Coulomb interactions with ions and electrons.
We use the expressions in \citet{Draine:1979a}, equations~(4)--(6),
considering the contributions from protons, electrons, helium
ions,\footnote{Helium is assumed to be singly ionized, leading to only
  a small contribution to the drag force.  For much hotter stars, such
  as the central stars of planetary nebulae, helium may be doubly
  ionized, which leads to a fourfold increase in its Coulomb drag
  contribution, which is significant for \(w < \SI{5}{km.s^{-1}}\).} %
and metal ions.\footnote{%
  \label{fn:metal-drag}
  All metals are lumped together as a single species, assuming
  standard \hii{} region gas-phase abundances.  They are dominated by
  C and O, with minor contributions from N and Ne.  The total
  abundance is \num{8.5e-4} and the effective atomic weight is
  \num{15.3}.  All are assumed to be doubly ionized.  Their largest
  relative contribution to the drag force is for
  \(w < \SI{2}{km.s^{-1}}\), but is less than 1\% even there.} %
Results are shown in Figure~\ref{fig:drag-components}, where dashed lines correspond to direct solid body collisions and solid lines to electrostatic interactions.  The latter depend on the grain potential, which is described in dimensionless terms by \(\phi\), which is the electrostatic potential
energy of a unit charge at the surface of a grain of charge
\(Z\grain\) and radius \(a\), in units of the characteristic thermal
energy of a gas particle:
\begin{equation}
  \label{eq:phi-potential}
  \phi = \frac{e^2 Z\grain}{a kT} \ .
\end{equation}
The electrostatic contributions to \(f\drag\) are proportional to
\(\phi^2\) (results are shown for \(\abs{\phi} = 10\)), whereas the
solid-body contributions are independent of \(\phi\).  The drag force is
put in dimensionless units by dividing by a characteristic force:
\begin{equation}
  \label{eq:fstar}
  f_* = 2 n k T \cdot \pi a^2 \ , 
\end{equation}
which is approximately\footnote{%
  To simplify the exposition, the gas pressure in this section is
  calculated assuming a fully ionized, pure hydrogen plasma, yielding
  \(P\gas = 2 n k T\). For typical ISM abundances, the contribution of
  helium and its corresponding electrons yield a correction to this of
  order 5\%.  The required modifications when a cool star interacts
  with a predominantly neutral gas stream are discussed later.  } %
the ionized gas pressure multiplied by the grain geometric cross
section.

For grains with low electric charge, \(\phi^2 \ll 1\), the drag force is
dominated by direct collisions of protons with the grain (dashed red
line in Fig.~\ref{fig:drag-components}).  The gas collisional mean
free path is much larger than the grain size, so the drag is in the
Epstein regime \citep{Weidenschilling:1977b}.
% This is illustrated by
% the \(\phi = 0.25\) case (blue line) in
% Figure~\ref{fig:gas-grain-drag-photoionized}.
As the relative gas--grain slip speed, \(w\), increases, \(f\drag\)
first increases linearly with \(w\) reaching \(f\drag \approx f_*\) at
\(w = \sound \approx \SI{10}{km.s^{-1}}\), then transitions to a quadratic
increase in the supersonic regime.

As \(\abs{\phi}\) increases, long-range electrostatic interactions with
protons within the Debye radius (Coulomb drag) become increasingly
important at subsonic relative velocities, as shown by the solid lines
in Figure~\ref{fig:drag-components}).
% in
% Figure~\ref{fig:gas-grain-drag-photoionized} by the orange
% (\(\phi = 1\)), green (\(\phi = 4\)), and red (\(\phi = 16\)) lines.
However, the Coulomb drag has a peak when \(w\) is equal to the
thermal speed of the colliders, which is
\(\approx \SI{10}{km.s^{-1}}\) for protons, giving a maximum strength of
\begin{equation}
  \label{eq:fdrag-maximum}
  f_{\mathrm{max}} = 0.5\, (\ln\Lambda)\, \phi^2 f_* \approx 10\, \phi^2 f_* \ , 
\end{equation}
where \(\Lambda\) is the plasma parameter (number of particles within a
Debye volume), such that
\(\ln\Lambda = 23.267 + 1.5 \ln T_4 - 0.5 \ln n\).  At highly super-thermal
speeds, the Coulomb drag falls asymptotically as
\(f\drag \propto 1 / w^{2}\).  The thermal speed of electrons is higher than
that of the protons by a factor of \((m_p / m_e)^{1/2}\), so that the
electron Coulomb drag (solid light blue line) gives a second peak of
similar strength, but at \(w \approx \SI{430}{km.s^{-1}}\).  The behavior of
\(f\drag\) in all these different regimes is summarised in
Table~\ref{tab:fdrag-regimes}, in terms of \(\phi\) and
\(w_{10} = w / \SI{10}{km.s^{-1}}\).  This is further illustrated in
Figure~\ref{fig:drag-v-phi-plane}, where each of the drag regimes is
located on the \((w, \abs{\phi})\) plane.

@@

From equations~\eqref{eq:dust-rad-force} and~\eqref{eq:fstar}, we can
write the radiation force acting on a grain as
\begin{equation}
  \label{eq:frad-Xi}
  f\rad = \Qp\, \Xi\, f_* \ ,
\end{equation}
where \(\Qp\) is the grain's radiation pressure efficiency (see
footnote~\ref{fn:Qp} in Appendix~\ref{sec:gas-free-bow}) and \(\Xi\) is
the local radiation parameter, defined as the ratio of direct stellar
radiation pressure to gas pressure:
\begin{equation}
  \label{eq:Xi-Prad-over-Pgas}
  \Xi \equiv \frac{P\rad}{P\gas} \approx \frac{L}{4 \pi R^2 c\, (2 n k T)} \ ,
\end{equation}
where the last expression corresponds to the optically thin limit.
The grain potential \(\phi\) is also primarily determined by \(\Xi\), as
shown in Appendix~\ref{sec:cloudy-models-dust} and the inset graph of
Figure~\ref{fig:gas-grain-drag-photoionized}, but with a slow
dependence, which can be approximated as
\begin{equation}
  \label{eq:phi-vs-Xi}
  \phi(\Xi) \approx 1.5 \bigl( 2.3 +  \ln \Xi \bigr) \ .
\end{equation}
There are also slight secondary dependencies on the grain composition and
stellar spectrum.  The relationship given in eq.~\eqref{eq:phi-vs-Xi}
is appropriate for graphite grains and for stellar effective
temperatures in the range \SIrange{20}{30}{kK}.  For hotter stars than
this, \(\phi\) should be multiplied by a further factor of \(1.5\), while
for silicate grains it should be divided by \(1.5\).

@@

\begin{table}
  \caption{Critical values of radiation parameter at the rip point: \(\Xi_\dag\)}
  \centering
  \begin{tabular*}{0.75\columnwidth}{l @{\quad\quad\quad\quad} S S} \toprule
    & \multicolumn{2}{c}{Grain composition} \\
    Spectrum & {Graphite} & {Silicate}
    \\ \midrule
    B star & 1000 +- 400 & 350 +- 150 \\
    O star & 3000 +- 500 & 2500 +- 500 \\
    \bottomrule
    \addlinespace
    \multicolumn{3}{@{}p{0.75\columnwidth}@{}}{
    Calculated from the Cloudy models shown in Figure~\ref{fig:drift-gn}. 
    Uncertainties represent variations with grain size and gas density. 
    See Appendix~\ref{sec:cloudy-models-dust} for further details.}
  \end{tabular*}
  \label{tab:Xi-rip}
\end{table}

As the dusty stream approaches the star, the radiation parameter
\(\Xi\) will increase, with a dependence of \(R^{-2}\) once the stream
is well inside the ionization front.  This increases \(f\rad\)
(eq.~\eqref{eq:frad-Xi}), but also increases the grain potential,
\(\phi\) (eq.~\eqref{eq:phi-vs-Xi}) due to the increasing dominance of
grain charging by photoelectric ejection.  Initially, this results in
a lowering of the equilibrium drift velocity to \(w_{10} \sim 0.01\) as
the Coulomb drag kicks in (see Appendix~\ref{sec:cloudy-models-dust}).
However, at smaller radii the slow logarithmic increase in
\(\phi(\Xi)\) means that the drift velocity must start increasing again to
accommodate the linear increase of \(f\rad(\Xi)\).  Eventually,
\(f\rad\) exceeds \(f_{\mathrm{max}}\), the maximum drag force that
proton Coulomb interactions can provide
(eq.~\eqref{eq:fdrag-maximum}).  This occurs at a critical value of
the radiation parameter, which we denote the \textit{rip point}:
\(\Xi_\dag \sim 1000\).  The variations in \(\Xi_\dag\) with star and grain
parameters, which are of order \SI{+- 0.5}{dex}, are listed in
Table~\ref{tab:Xi-rip} and illustrated graphically in
Figure~\ref{fig:drift-gn}.  The radius of the rip point, \(R_\dag\), can
be expressed in terms of \(R_*\), the fiducial optically thick bow
shock radius introduced in \S~\ref{sec:strong-gas-grain}:
\begin{equation}
  \label{eq:Rdag-over-Rstar}
  R_\dag = \frac{v_\infty}{\sound}\, \Xi_\dag^{-1/2} R_* \approx v_{10}\, \Xi_\dag^{-1/2} R_* \ ,
\end{equation}
where we have made use of equations~\eqref{eq:Rstar} and~\eqref{eq:Xi-Prad-over-Pgas}.

\subsubsection{Grain trajectories for the on-axis case}
\label{sec:grain-traj-axis}

What happens to the dust grain following this catastrophic breakdown
of gas--grain coupling depends on the relation between the rip point
radius, \(R_\dag\), and the drag-free radiative turnaround radius,
\(R\starstar\).  If \(R_\dag > R\starstar\), then the grain's inertia
will still carry it in as far as \(R\starstar\) and the initial
trajectory will be almost identical to that described in
Appendix~\ref{sec:gas-free-bow} for the drag-free case. However, after
being turned around by the radiation field and pushed out past
\(R_\dag\) again, the grain will \emph{recouple} to the gas and be
dragged back for a second approach.\footnote{If the initial impact
  parameter of the trajectory is not strictly \(b = 0\), then the
  resultant lateral component of \(f\rad\) will mean that \(b\) will
  be much increased for the second approach.} %
If, on the other hand, \(R_\dag < R\starstar\), then the tail wind
provided by the gas drags the grain closer to the star than it
``wants'' to be.

\subsubsection{Dust waves in stream parameter space}
\label{sec:dust-waves-stream}

\section{The case of inside-out bows}
\label{sec:case-inside-out}
